|

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): note nameref option for LaTeX outputs
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(manuscript): fix suppfig references vs caption titles
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(manuscript): update text
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@

      reference-prefix: Fig. S
      space-before-numbering: false

@@

>>>>

# Use of `\nameref{sec-methods}` etc. to refer to sections by name.
# is only compatible with LaTeX outputs

@@

      reference-prefix: Supplementary Fig.
      caption-prefix: Supplementary Figure
      space-before-numbering: true

@@

One major biological hurdle for current formulations of RNA Velocity models has
been miscalling cell trajectories in fully differentiated cell populations that
lack dynamic cell state transitions. One typical example being human peripheral
blood mononuclear cells (PBMCs) [@Zheng2017-bz]. In this dataset, 65,877 cells
were analyzed using the 10x platform that comprised 11 fully differentiated,
mature immune cell types. This dataset lacked stem and progenitor cells or other
signatures of an undergoing dynamical differentiation; thus, no consistent
velocity flow should be detected. We reanalyzed this dataset using scVelo and
Pyro-Velocity using a common set of genes for a fair comparison. Using scVelo,
we reproduced the original results showing incorrect cell trajectory assignments
([@fig-model-results]a) [@Bergen2021-qz]. ScVelo uncovered an errant, strong
linear trajectory across a subset of mature cell types [@Zheng2017-bz] that is
exasperated by the inability to confidently assess lineage directionality or a
baseline to assess significance. On the contrary, Pyro-Velocity failed to detect
high-confidence trajectories in the mature blood cell states, consistent with
what is known about the biology underlying these cells ([@fig-model-results]a,
panel 2 vs. panel 3). Importantly, the Bayesian nature of Pyro-Velocity enabled
the generation of uncertainty estimates that can be easily visualized as likely
vectors for individual cells or averaged across cells on a constantly spaced
grid on the cell embedding (*averaged vector field*).

Pyro-Velocity was also able to define well-known developmental cell hierarchies
in a scRNA-seq dataset of E15.5 mouse pancreas ([@fig-model-results]b-e,
$n=3,696$ cells comprising $8$ cell types ([@fig-model-results]b, panel 1)),
identifying cell trajectories originating from ductal progenitor cells and
culminated in the production of mature Alpha, Beta, Delta, and Epsilon cells
[@Bastidas-Ponce2019-lf]. Though the scVelo and Pyro-Velocity models have
qualitatively comparable vector fields, the results generated by scVelo were
greatly influenced by gene selection, neighboring cell number for gene
expression smoothing, and cannot provide confidence assessment of lineage
directionality ([@suppfig-model-results]a). In addition, Pyro-Velocity
uncertainty visualizations uncovered that differentiated cells had a larger
vector field uncertainty than stem and progenitor cells, as would be expected
for terminally differentiated cell states where progression and lineage
trajectories would not be anticipated. Pyro-Velocity also recovered a
developmental order in close agreement with what was estimated by Cytotrace
(Spearman’s $r=0.95, \; \text{p-value} < 10^{-10}$ between latent time and
developmental order)[@Gulati2020-xq] - a state-of-the-art orthogonal method used
to predict cell differentiation based on the number of expressed genes per cell.
